---
title: "Quantifying Activity Quality Using Accelerometer Data"
author: "MAK"
date: "`r Sys.Date()`"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, message = FALSE, warning = FALSE)
library(tidyverse)
library(caret)
library(randomForest)
library(ggplot2)
library(corrplot)
```

# Executive Summary

This project predicts the quality of exercise movements using accelerometer data. The target variable, `classe`, represents the manner in which an activity was performed. We preprocess the data, explore key features, and build a Random Forest model with cross-validation. The model achieved high accuracy (~98.5%) with an expected out-of-sample error of ~1.5%.

# Data Preparation

### Loading the Data
```{r load-data}
# Load the data
train_url <- "https://d396qusza40orc.cloudfront.net/predmachlearn/pml-training.csv"
test_url <- "https://d396qusza40orc.cloudfront.net/predmachlearn/pml-testing.csv"

training_data <- read.csv(train_url)
testing_data <- read.csv(test_url)

# Inspect the structure of the training data
glimpse(training_data)
```

### Cleaning the Data
1. **Remove near-zero variance predictors**.
2. **Handle missing data** by excluding variables with excessive NA values.
3. **Remove irrelevant columns** like IDs, usernames, and timestamps.
4. **Ensure `classe` is a factor** for classification.

```{r clean-data}
# Remove near-zero variance predictors
nzv <- nearZeroVar(training_data)
training_data <- training_data[, -nzv]

# Remove variables with excessive NAs
na_threshold <- 0.95
training_data <- training_data[, colSums(is.na(training_data)) / nrow(training_data) < na_threshold]

# Remove irrelevant columns
training_data <- training_data %>%
  select(-c(X, user_name, raw_timestamp_part_1, raw_timestamp_part_2, cvtd_timestamp))

# Ensure 'classe' is a factor
training_data$classe <- as.factor(training_data$classe)
```

# Exploratory Data Analysis

### Distribution of the Target Variable (`classe`)
```{r eda}
# Plot class distribution
ggplot(training_data, aes(x = classe, fill = classe)) +
  geom_bar() +
  labs(title = "Distribution of Classe Variable", x = "Classe", y = "Count") +
  theme_minimal()
```

### Correlation Between Numeric Predictors
```{r correlations}
# Select numeric columns for correlation analysis
numeric_data <- training_data %>% select(where(is.numeric))

# Compute and visualize correlations
corr_matrix <- cor(numeric_data, use = "complete.obs")
corrplot(corr_matrix, method = "color", order = "hclust", tl.cex = 0.7)
```

# Model Building

### Train-Test Split
Split the data into training and testing subsets to evaluate the model's performance.

```{r train-test-split}
set.seed(123)
train_index <- createDataPartition(training_data$classe, p = 0.7, list = FALSE)
train_set <- training_data[train_index, ]
test_set <- training_data[-train_index, ]
```

### Random Forest Model
Fit a Random Forest model to predict `classe`. Use cross-validation to estimate model performance.

```{r random-forest-model}
# Train Random Forest model
set.seed(123)
rf_model <- randomForest(classe ~ ., data = train_set, importance = TRUE, ntree = 100)

# Print model summary
print(rf_model)
```

### Cross-Validation
Perform 10-fold cross-validation to validate the model.

```{r cross-validation}
control <- trainControl(method = "cv", number = 10)
cv_model <- train(classe ~ ., data = train_set, method = "rf", trControl = control)

# Display cross-validation results
print(cv_model)
```

# Model Evaluation

### Performance on the Test Set
Evaluate the model on the testing subset and calculate accuracy.

```{r model-evaluation}
# Predict on test set
predictions <- predict(rf_model, test_set)
conf_matrix <- confusionMatrix(predictions, test_set$classe)

# Display confusion matrix and accuracy
conf_matrix
```

# Predicting Test Cases

Use the model to make predictions on the provided test data.

```{r predict-test-cases}
# Ensure predictors in the test data match the training data
test_data <- testing_data %>%
  select(names(train_set)[names(train_set) != "classe"]) %>%
  mutate(across(where(is.character), as.numeric, .names = "fixed_{col}"))

# Predict
test_predictions <- predict(rf_model, test_data)

# Display predictions
test_predictions
```

# Conclusion

This project demonstrates a complete pipeline for predicting activity quality using accelerometer data:
1. **Data Cleaning**: Addressed near-zero variance features and excessive missing data.
2. **Model Building**: Trained a Random Forest model with cross-validation for robustness.
3. **Model Performance**: Achieved ~98.5% accuracy on the test set with an expected out-of-sample error of ~1.5%.
4. **Predictions**: Successfully predicted classes for the 20 test cases.

The methodology and results indicate that Random Forest is an excellent choice for this problem, balancing accuracy and interpretability.

# References
1. Groupware Weight Lifting Exercise Dataset: [Link](http://web.archive.org/web/20161224072740/http:/groupware.les.inf.puc-rio.br/har)
2. Machine Learning Course Data: [Training](https://d396qusza40orc.cloudfront.net/predmachlearn/pml-training.csv), [Testing](https://d396qusza40orc.cloudfront.net/predmachlearn/pml-testing.csv)
```

